Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/semantics.tex
revise proof of lemma 4

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole{}, which represent \emph{unconstrained}
values that may replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, where they may be thought
of as type variables that we will not generalize over.
%

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{smallcheck}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes (denoted by their index)
to values which may, in turn, contain holes.
%
The substitutions let us ensure that we instantiate each hole
with the same (partially defined) value, regardless of the multiple
contexts in which the hole appears, and hence, ensures we can
report a concrete (and general) witness for any (dynamically)
discovered type errors.

@@

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
yields the stuck term if no such refinement is possible. In the case
where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
if the hole has already been instantiated and, if so, returns the
existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Steps and Traces}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
$\pair{e_0}{\vsu_0}, \ldots, \pair{e_n}{\vsu_n}$ such that
$\forall 0 \leq i < n$, we have \step{e_i}{\vsu_i}{e_{i+1}}{\vsu_{i+1}}.
%
We write \steptr{\trace}{e}{\vsu}{e'}{\vsu'} if $\trace$ is a trace of the form
$\pair{e}{\vsu},\ldots,\pair{e'}{\vsu'}$.
%
We write \steps{e}{\vsu}{e'}{\vsu'} if \steps{e}{\vsu}{e'}{\vsu'}
for some trace $\trace$.

\paragraph{Primitive Reductions}
Primitive reduction steps -- addition, if-elimination, and
function application -- use \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
step extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k+1$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
     \tincompat{t}{\ptype{\trace}{f}}, and
     \hastype{v}{t},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\ES{self-todo: take a close look at the $\trace$'s vs k's}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k+1$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace}}:]
    The inductive hypothesis applies.
  %\item
  [Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  %\end{description}
\end{proof}

\paragraph{Proof of Theorem~\ref{thm:soundness}}
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t$ be the partial type of $f$ upto $\trace$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with the partial type upto $\trace$.
%
% \begin{description}
  %\item
  [Case \tcompat{s}{t}:]
   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
   gets stuck.
  %\item
  [Case \tincompat{s}{t}:]
   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
   $\eapp{f}{v}$ gets stuck.
%\end{description}

>>>>

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole{}, which represents \emph{unconstrained}
values that may replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, where they may be thought
of as type variables that we will not generalize over.
%

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes (denoted by their index)
to values which may, in turn, contain holes.
%
The substitutions let us ensure that we instantiate each hole
with the same (partially defined) value, regardless of the multiple
contexts in which the hole appears, and hence, ensures we can
report a concrete (and general) witness for any (dynamically)
discovered type errors.

@@

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and an empty value
substitution, or yields the stuck term if no such refinement is
possible. In the case where \texttt{e} is a hole, it first checks in the
given $\vsu$ to see if the hole has already been instantiated and, if
so, returns the existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Steps and Traces}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
$\pair{e_0}{\vsu_0}, \ldots, \pair{e_n}{\vsu_n}$ such that
$\forall 0 \leq i < n$, we have \step{e_i}{\vsu_i}{e_{i+1}}{\vsu_{i+1}}.
%
We write \steptr{\trace}{e}{\vsu}{e'}{\vsu'} if $\trace$ is a trace of the form
$\pair{e}{\vsu},\ldots,\pair{e'}{\vsu'}$.
%
We write \steps{e}{\vsu}{e'}{\vsu'} if \steptr{\trace}{e}{\vsu}{e'}{\vsu'}
for some trace $\trace$.

\paragraph{Primitive Reductions}
Primitive reduction steps --- addition, if-elimination, and
function application --- use \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
% \ES{self-todo: take a close look at the $\trace$'s vs k's}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$ and
  $\trace' = \pair{\eapp{f}{\ehole{}}}{\emptysu},...,\pair{e'}{\vsu'}$ 
  such that $\step{e'}{\vsu'}{e}{\vsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  %\item
  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  %\end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
% \begin{description}
  %\item
  [Case \tcompat{s}{t}:]
   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
   gets stuck.
  %\item
  [Case \tincompat{s}{t}:]
   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
   $\eapp{f}{v}$ gets stuck.
%\end{description}
\end{proof}
